Highly Available Transactions: Virtues and Limitations

### 2013-05-01 writeup/motivation.tex
pass, plus spelling and grammar

Why does availability matter? Peter Deutsch starts his classic list of
``Fallacies of Distributed Computing'' with two concerns fundamental
to distributed database systems: ``\textit{1.)}  The network is
reliable. \textit{2.)} Latency is zero''~\cite{fallacies-deutsch}; in
a distributed setting, network failures may prevent database servers
from communicating, and, in the absence of failures, communication is
slowed by factors like physical distance and, more often, congestion
and routing. As we will see (Section~\ref{sec:availability}), highly
available system operation minimizes the effect of network
\textit{partitions} and \textit{latency}. In this Section, we draw on
a range of evidence that indicates that partitions occur relatively
frequently in real-world deployments and best-case latencies between
datacenters are substantial, often on the order of hundreds of
milliseconds.

@@

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Services team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 outage in October 2012 have led
to full site disruptions for popular web services like Reddit,
Foursquare, and Heroku~\cite{ec2-downsites}. At a global scale,
hardware failures---like the 2011 outages in Internet backbones in
North America and Europe due a router
bug~\cite{juniper-partition}---and misconfigurations like the BGP
faults in 2008~\cite{pakistan-youtube} and
2010~\cite{research-experiment-partition} can cause widespread
partitions.

@@

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' In this
section, we quantify round-trip latencies, which are often
large---hundreds of milliseconds---in a geo-replicated,
multi-datacenter context.  Fundamentally, the speed at which two
servers can communicate is (according to modern physics) bounded by
the speed of light. In the best case, two servers on opposite sides of
the Earth---communicating via a hypothetical link along the planet's
diameter---require a minimum 85.1ms RTT (133.7ms if sent at
surface level). As services are replicated to multiple, geographically
distinct sites, this cost of communication increases.

@@

\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
IR & & & &  & 346.2   & 239.8   & 234.1  \\
SY & & & & &  & 333.6   & 243.1  \\
SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
\hline
  \end{tabular}
}

\caption{Mean RTT times on EC2 (min and max highlighted)}\vspace{-1em}
\label{table:rtt}
\end{table}

@@

In real deployments, messages travel slower than the speed of light
due to routing, congestion, and server-side overheads. To illustrate
the difference between intra-datacenter, inter-datacenter, and
inter-planetary networks, we performed a measurement study of network
behavior on Amazon's EC2, a widely-used public compute cloud. We
measured one week of ping times between all seven EC2 geographic
``regions'', across three ``availability zones'' (closely co-located
datacenters), and within a single ``availabilty zone'' (datacenter),
at a granularity of 1s (dataset to be released). We summarize the
results of our network measurement study in Table~\ref{table:rtt}. On
average, intra-datacenter communication (Table 1a) is between 1.82 and
6.38 times faster than across geographically co-located datacenters
(Table 1b) and between 40 and 647 times faster than across
geographically distributed datacenters (Table 1c). The cost of
wide-area communication exceeds the speed of light: for example, while
a speed-of-light RTT from S\~{a}o Paulo to Singapore RTT is 106.7ms,
ping packets incur an average 362.8ms RTT (95th percentile: 649ms). As
shown in Figure~\ref{fig:rtt}, the distribution of latencies varies
between links, but the trend is clear: remote communication has a
substantial cost.

@@

The previous section demonstrated that distributed systems must
address partitions and latency: what does this mean for distributed
databases? Database researchers and designers have long realized that
serializability is not achievable in a highly available
system~\cite{davidson-survey}, meaning that, in environments like
those in Section~\ref{sec:motivation}, database designs face a choice
between availability and strong semantics. However, even in a
single-node database, the coordination penalties associated with
serializability can be severe, manifesting themselves in the form of
decreased concurrency (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Accordingly, to increase
concurrency, database systems offer a range of ACID properties weaker
than serializability.  Instead, databases offer a host of so-called
\textit{weak isolation} models that allow varying restrictions on the
space of schedules that are allowable by the system~\cite{adya,
  ansicritique, ansi-sql}. None of these weak isolation models
guarantees serializability, but, as we see below, their benefits are
often considered to outweigh costs of possible consistency anomalies
that might arise from their use.

To understand the prevalence of these weak isolation models, we
recently surveyed the default and maximum isolation models provided by
18 databases, often claiming to provide ``ACID'' or ``NewSQL''
functionality~\cite{hat-hotos}. As shown in
Table~\ref{table:existing}, only three out of 18 databases provided
serializability by default, and eight did not provide serializability
as an option at all. This is particularly surprising when we consider
the widespread deployment of many of these non-serializable databases,
like Oracle 11g, which are known to power major businesses and product
functionality. Given that these transactional models are frequently
used, our inability to provide serializability in arbitrary HATs
appears non-fatal for practical applications. If application writers
and database vendors have already decided that the benefits of weak
isolation outweigh potential application inconsistencies, then, in a
highly available environment that prohibits serializability, similar
decisions may be tenable.

@@

It is unknown \textit{which} of these guarantees can be provided with
high availability. The existing algorithms for providing weak
isolation are typically designed for a single-node context and are, to
the best of our knowledge, unavailable due to a reliance on
concurrency control mechanisms like locking that are not resilient to
partial failure (Section~\ref{sec:evaluation}). Moreover, we are not
aware of any prior literature that provides guidance as to the
relationship between weak isolation and high availability: prior work
has examined the relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a single
server~\cite{adya, ansicritique, gray-isolation} but never weak
isolation and high availability taken together.  A primary goal in the
remainder of this paper is to understand which models are
HAT-compliant.

>>>>

Why does high availability matter? Peter Deutsch starts his classic
list of ``Fallacies of Distributed Computing'' with two concerns
fundamental to distributed database systems: ``\textit{1.)}  The
network is reliable. \textit{2.)} Latency is
zero''~\cite{fallacies-deutsch}; in a distributed setting, network
failures may prevent database servers from communicating, and, in the
absence of failures, communication is slowed by factors like physical
distance and, more often, congestion and routing. As we will see
(Section~\ref{sec:availability}), highly available system design
mitigates the effects of network \textit{partitions} and
\textit{latency}. In this section, we draw on a range of evidence that
indicates that partitions occur relatively frequently in real-world
deployments and best-case latencies between datacenters are
substantial, often on the order of several hundreds of milliseconds.

@@

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Services team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 outage in October 2012 have led
to full site disruptions for popular web services like Reddit,
Foursquare, and Heroku~\cite{ec2-downsites}. At global scale,
hardware failures---like the 2011 outages in Internet backbones in
North America and Europe due a router
bug~\cite{juniper-partition}---and misconfigurations like the BGP
faults in 2008~\cite{pakistan-youtube} and
2010~\cite{research-experiment-partition} can cause widespread
partitions.

@@

Even with fault-free networks, distributed systems face the challenge
of network communication latency, Deutsch's second ``Fallacy.'' In this
section, we quantify round-trip latencies, which are often
large---hundreds of milliseconds in a geo-replicated,
multi-datacenter context.  Fundamentally, the speed at which two
servers can communicate is (according to modern physics) bounded by
the speed of light. In the best case, two servers on opposite sides of
the Earth---communicating via a hypothetical link along the planet's
diameter---require a minimum 85.1ms RTT (133.7ms if sent at
surface level). As services are replicated to multiple, geographically
distinct sites, this cost of communication increases.

@@

\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)\vspace{-.5em}] {
  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
IR & & & &  & 346.2   & 239.8   & 234.1  \\
SY & & & & &  & 333.6   & 243.1  \\
SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
\hline
  \end{tabular}
}

\caption{Mean RTT times on EC2 (min and max highlighted)}\vspace{-.5em}
\label{table:rtt}
\end{table}

@@

In real deployments, messages travel slower than the speed of light
due to routing, congestion, and server-side overheads. To illustrate
the difference between intra-datacenter, inter-datacenter, and
inter-planetary networks, we performed a measurement study of network
behavior on Amazon's EC2, a widely used public compute cloud. We
measured one week of ping times (i.e., round-trip times, or
RTTs) between all seven EC2 geographic ``regions,'' across three
``availability zones'' (closely co-located datacenters), and within a
single ``availability zone'' (datacenter), at a granularity of 1s
(dataset to be released). We summarize the results of our network
measurement study in Table~\ref{table:rtt}. On average,
intra-datacenter communication (Table 1a) is between 1.82 and 6.38
times faster than across geographically co-located datacenters (Table
1b) and between 40 and 647 times faster than across geographically
distributed datacenters (Table 1c). The cost of wide-area
communication exceeds the speed of light: for example, while a
speed-of-light RTT from S\~{a}o Paulo to Singapore RTT is 106.7ms,
ping packets incur an average 362.8ms RTT (95th percentile: 649ms). As
shown in Figure~\ref{fig:rtt}, the distribution of latencies varies
between links, but the trend is clear: remote communication has a
substantial cost.

@@

The previous section demonstrated that distributed systems must
address partitions and latency: what does this mean for distributed
databases? Database researchers and designers have long realized that
serializability is not achievable in a highly available
system~\cite{davidson-survey}, meaning that, in environments like
those in Section~\ref{sec:motivation}, database designs face a choice
between availability and strong semantics. However, even in a
single-node database, the coordination penalties associated with
serializability can be severe, manifesting themselves in the form of
decreased concurrency (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Accordingly, to increase
concurrency, database systems offer a range of ACID properties weaker
than serializability: the host of so-called \textit{weak isolation}
models describe varying restrictions on the space of schedules that
are allowable by the system~\cite{adya, ansi-sql, ansicritique}. None
of these weak isolation models guarantees serializability, but, as we see below,
their benefits are often considered to outweigh costs of possible
consistency anomalies that might arise from their use.

To understand the prevalence of weak isolation, we
recently surveyed the default and maximum isolation guarantees provided by
18 databases, often claiming to provide ``ACID'' or ``NewSQL''
functionality~\cite{hat-hotos}. As shown in
Table~\ref{table:existing}, only three out of 18 databases provided
serializability by default, and eight did not provide serializability
as an option at all. This is particularly surprising when we consider
the widespread deployment of many of these non-serializable databases,
like Oracle 11g, which are known to power major businesses and product
functionality. Given that these transactional models are frequently
used, our inability to provide serializability in arbitrary HATs
appears non-fatal for practical applications. If application writers
and database vendors have already decided that the benefits of weak
isolation outweigh potential application inconsistencies, then, in a
highly available environment that prohibits serializability, similar
decisions may be tenable.

@@

It is unknown \textit{which} of these guarantees can be provided with
high availability, or are HAT-compliant. Existing algorithms for
providing weak isolation are often designed for a single-node context
and are, to the best of our knowledge, unavailable due to reliance
on concurrency control mechanisms like locking that are not resilient
to partial failure (Section~\ref{sec:eval-existing}). Moreover, we are
not aware of any prior literature that provides guidance as to the
relationship between weak isolation and high availability: prior work
has examined the relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation in
general~\cite{adya, ansicritique, gray-isolation} but not weak
isolation and high availability together.  A primary goal in the
remainder of this paper is to understand which models are
HAT-compliant.

### 2013-05-01 writeup/relatedwork.tex
PDB pass
pass, plus spelling and grammar

There is a growing body related work on highly available semantics,
mechanisms for concurrency control, and techniques for scalable
distributed operations.

Weak consistency and high availability have been well
studied. Serializability has long been known to be
unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
attracted considerable attention~\cite{gilbert-cap}. Recent work on
PACELC expands CAP by considering connections between ``weak
consistency'' and low latency~\cite{abadi-pacelc}, while several
studies examine weak isolation guarantees~\cite{adya,
  ansicritique}. However, as we have discussed, the connection between
weak, multi-object consistency models in a distributed setting and
high availability is relatively unexplored. With the exception of our
preliminary workshop paper discussing transactional availability and
HAT RC and I-CI~\cite{hat-hotos}--- which this work expands, providing
additional algorithms and impossibility results, a full
taxonomization, and experimental and application analysis---we believe
this paper is the first to explore this area. However, several studies
rigorously classify \textit{non-HAT} semantics: Wisemann et al. have
proposed a three parameter classification for one-copy serializable
database systems, performing a more extensive classification than our
overview in Section~\ref{sec:evaluation}, including eager database
replication techniques~\cite{kemme-classification} and distributed
mechanisms for achieving one-copy
serializability~\cite{wisemann-survey}.

@@

With HATs, we seek an alternative set of semantics that are still
useful but do not violate requirements for high availability or low
latency. Recent systems proposals such as Swift~\cite{swift},
Eiger~\cite{eiger}, and Bolt-on Causal Consistency~\cite{bolton}
provide transactional causal consistency guarantees with varying
availability and represent a new class of sticky HAT systems. There
are infinitely many HAT models (i.e., always reading value 1 is
comparable with always returning value 2), but a recent report from UT
Austin shows that no model stronger than causal consistency can be
achieved in a sticky highly available, \textit{one-way convergent}
system~\cite{cac}. This result is promising and complementary to our
results for general-purpose convergent data stores.

>>>>

While we have discussed traditional mechanisms for distributed
coordination and several related systems in
Section~\ref{sec:eval-existing}, in this section, we further discuss
related work. In particular, we discuss related work on highly
available semantics, mechanisms for concurrency control, and
techniques for scalable distributed operations.

Weak consistency and high availability have been well
studied. Serializability has long been known to be
unachievable~\cite{davidson-survey} and Brewer's CAP Theorem has
attracted considerable attention~\cite{gilbert-cap}. Recent work on
PACELC expands CAP by considering connections between ``weak
consistency'' and low latency~\cite{abadi-pacelc}, while several
studies examine weak isolation guarantees~\cite{adya,
  ansicritique}. However, as we have discussed, the connection between
weak, multi-object consistency models in a distributed setting and
high availability is relatively unexplored. With the exception of our
preliminary workshop paper discussing transactional availability,
real-world ACID, and HAT RC and I-CI~\cite{hat-hotos}---which this
work expands, providing additional algorithms and impossibility
results, a full taxonomization, experimental and application analysis,
and much greater discussion of implications and related work---we
believe this paper is the first to explore this area. However, several
studies rigorously classify \textit{non-HAT} semantics: Wisemann et
al. have proposed a three parameter classification for one-copy
serializable database systems, performing a more extensive
classification than our overview in Section~\ref{sec:evaluation},
including eager database replication
techniques~\cite{kemme-classification} and distributed mechanisms for
achieving one-copy serializability~\cite{wisemann-survey}.

@@

With HATs, we seek an alternative set of semantics that are still
useful but do not violate requirements for high availability or low
latency. Recent systems proposals such as Swift~\cite{swift},
Eiger~\cite{eiger}, and Bolt-on Causal Consistency~\cite{bolton}
provide transactional causal consistency guarantees with varying
availability and represent a new class of sticky HAT systems. There
are infinitely many HAT models (i.e., always reading value 1 is
comparable with always returning value 2), but a recent report from UT
Austin shows that no model stronger than causal consistency can be
achieved in a sticky highly available, \textit{one-way convergent}
system~\cite{cac}. This result is promising and complementary to our
results for general-purpose convergent data stores. Finally, Burkhardt
et al. have concurrently developed an axiomatic specification for
eventually consistent stores; their work-in-progress report contains
alternate formalism for several HAT guarantees and (snapshot)
transactions~\cite{burkhardt-txns}.
